 

### 2021-04-22 vertical-thermal-review.Rmd
Merge branch 'master' of https://github.com/EcoClimLab/vertical-thermal-review

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [Kenzo et al. 2015; Sanches et al. 2009]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [Sanches et al. 2009; Zweinniecki et al. 2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves; therefore decreases with height [Sanches et al. 2009; Hernandez]. Similarly, max transpiration rate increases with canopy height [Kenzo et al. 2015]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [Kosugi et al. 2021].

@@

**Leaf thermal tolerance (Tcrit/ T50)**
*(define)*
Typical values ~45-50$^\circ$C, varying somewhat across latitude/climate (O' sullivan et al. 2017), with elevation [@feeley_thermal_2020;  Slot et al unpublished data] and with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012]. Tcrit is greater for buffered leaves that maintain greater effective leaf with [Slot et al. 2021]
Perez and Feeley et al. 2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins. These species may thus be at a greater risk of thermal damage.  
[- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]

Supporting the significant negative relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated with leaf thermal tolerance in order to maximize net carbon assimilation. However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor does it increase carbon assimilation at higher temperature. Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintenance, thus not contributing to increased C assimilation. Species exhibiting low carbon assimilation rates tended to have greater stress tolerance. In addition, species with low stomatal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimilation before thermal limits are reached [@perezPSIIHeatTolerances2020]

**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, both measured near ground level (Slot et al. 2019). 
However, for *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher climatic heat stress and thus $T_{50}$ was observed in the lower north-facing canopy positions (Curtis et al. 2019). 
We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?). I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot

increased irradiance = increased VPD = increased air temp = midday stomatal closure =  decreased latent heat loss = increase leaf temperature = decreased C assimilation = greater PS2 heat tolerance = optimum rate of photosynthesis decreases = exhibiting lower t opt (optimum temp for c assimilation)

incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [@perezPSIIHeatTolerances2020]

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Conflicting $Q_10$ findings shows no definite trend.
Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
- Elsa: and see Perez, T.M., Socha, A., Tserej, O. and Feeley, K.J., 2020. PSII heat tolerances characterize thermal generalists and the upper limit of carbon assimilation. Plant, Cell & Environment.

>>>>

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [Kenzo et al. 2015; Sanches et al. 2009]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [Sanches et al. 2009; Zweinniecki et al. 2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [Sanches et al. 2009; Hernandez]. Similarly, max transpiration rate increases with canopy height [Kenzo et al. 2015]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [Kosugi et al. 2021].

@@

**Leaf thermal tolerance (Tcrit/ T50)**
$T_{crit}$ is the temperature at which efficiency of photosystem II starts to decrease, based on chlorphyll a flourescence; and $T_{50}$ represents temperature at which the efficiency of photosystem II had decreased by 50% [Slot et al. 2020]. Leaf thermal thresholds strongly reflect species adaptation to their microenvironment rather than macroclimate conditions [@feeley_thermal_2020; Slot et al. 2020; Perez and Feely 2020a] and phylogeny [Slot et al. 2020; Perez and Feeley 2020a].

Typical values across latitude and climate are ~45-50$^\circ$C (O' sullivan et al. 2017). Across species, in Panama, Columbia and across latitudes, $T_{50}$ and $T_{crit}$ decreased with elevation [@feeley_thermal_2020;  Slot et al 2020; O' sullivan et al. 2017] and varies with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012; Slot et al. 2020] and their thermal time constant. Species with large leaves, and slow heating and cooling time [higher thermal time constant] showed greater $T_{crit}$. On the other hand, leaves of species with greater construction costs, observed through LMA, had greater $T_{50}$ compared to species with low LMA [Slot et al. 2020]. However, higher photosynthetic heat tolerance does not necessary increase leaf thermal safety margins, rather, in across species sun leaves, higher photosynthetic heat tolerance experienced higher maximum temperatures, low optimum temperature for C assimilation [$T_{opt}$] and had narrower thermal safety margins [Perez and Feeley, 2020]

**Thermal tolerance (e.g., $T_{50}$) varies with exposure, but there are no studies isolating effect of height.**
Thermal tolerance $T_{crit}$ and $T_{50}$ have been predominantly measured on sun leaves; thus there remains limited information on leaves with varying light exposures and vertical height gradient.

For two of three species studied in a tropical moist forest in Panama, $T_{50}$ was slightly lower for shade than sun-exposed leaves, though the results weren't significant and both were measured near ground level [Slot et al. 2019]. In tall grass prarie gallery forest in Kansas, photosynthetic heat tolerance measured on chinquapin oak (*Quercus muehlenbergii*) that grew in xeric, sunnier conditions had higher heat tolerance (by $2{^\circ}C$) than bur oak (*Quercus macrocarpa*) that grew in shaded mesic conditions. In both species photosynthetic heat tolerance increased from June to August.
In other cases, leaf thermal tolerance might reflect microclimatic stress along a vertical gradient. For *Acacia Papyrocarpa* (Benth.) trees in an arid region of southern Australia, higher composite climate stress and thus $T_{50}$ was observed in the lower north-facing canopy position compared to other canopy positions (Curtis et al. 2019). Greater $T_{50}$ in lower north-position correlated with longer leaf thermal time constants, along with with low-wind speed and greater canopy and air temperature [Curtis et al. 2018].

**We're not aware of any other studies on this."we may want to argue that we have no *a priori* reasons to expect thermal tolerance to scale with height per se (or do we?). I think it makes sense to expect variation with light exposure because of the associated differences in temperature you can expect the leaves to be exposed to, but height itself should not affect heat tolerance other than through associated traits perhaps"--Martijn Slot**

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Conflicting $Q_10$ findings shows no definite trend.
Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
